Temporal ordering and registration of images in studies of developmental dynamics

### 2015-01-26 image_analysis_paper_development2/manuscript.tex
updated paper

%\begin{figure}[t]
%\includegraphics{fig1}
%\caption{Caricature illustrating the tasks of image registration and temporal ordering. (A) Images of ``samples'', each in a different orientation and a different stage of development. (B) Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
%%\label{fig:fish}
%\customlabel{fig:fish}{1}
%\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
%\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
%\end{figure}

@@

%\begin{figure}[t]
%\includegraphics{fig2}
%\caption{Schematic illustrating angular synchronization and diffusion maps. (A) Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in A: the differences between a pair of angles in B is the same as the pairwise angle in A. (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \SI), so that close data points are connected by darker (``stronger'') edges. (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
%%\label{fig:schematics}
%\customlabel{fig:schematics}{2}
%\customlabel{subfig:synch1}{\ref{fig:schematics}A}
%\customlabel{subfig:synch2}{\ref{fig:schematics}B}
%\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
%\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
%\end{figure}

@@

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.75cm}{
{ \textsf{ 
\begin{tabular}{ |p{1cm} | p{1.25cm}|  p{1.5cm}| }
\hline
Movie Index & Average Angle Error & Rank Correlation Coefficient \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\ 
2 & 4.98$^{\circ}$ & 0.9994 \\ 
3 & 6.47$^{\circ}$ & 0.9795 \\ 
4 & 7.03$^{\circ}$ & 0.9953 \\ 
5 & 3.04$^{\circ}$ & 0.9959 \\ 
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies. }
\end{figure*}

\begin{figure*}[t]
\begin{minipage}[b]{14cm}
\raisebox{0.75cm}{\textsf{A}}
\includegraphics[width=13.75cm]{zebrafish_scrambled}

\raisebox{0.75cm}{\textsf{B}}
\includegraphics[width=13.75cm]{zebrafish_ordered}
\end{minipage}
%
\hfill
%
\raisebox{2.5cm}{\textsf{C}}
\includegraphics[width=3.25cm]{zebrafish_rank_corr}

\raisebox{0.5cm}{\textsf{C}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
\end{minipage}
%
\hfill
%
\raisebox{3.25cm}{\textsf{D}}
\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}

\begin{figure*}
\includegraphics[width=\textwidth]{wing_disc_ordered}
\end{figure*}

%\begin{figure*}[t]
%\includegraphics{fig3}
%\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies. (F) The median rank correlation coefficient (over $200$ samples) as a function of the number of images in the data set. }
%\customlabel{fig:movie}{3}
%\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}A}
%\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{B}}
%\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{C}}
%\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{D}}
%\customlabel{subfig:movie_error_table}{\ref{fig:movie}{E}}
%\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{F}}
%\end{figure*}

@@

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
%
Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%
To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%
Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%
\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%
As expected, the accuracy in the recovered ordering increases with the number of images. 
% 
At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
%
In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

We have analyzed how our algorithm performs on a model data set, where all images come from a single embryo. 
%
In practice, each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting an image from a random live imaging experiment at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.81, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.

@@

%\begin{figure*}[t]
%\includegraphics{fig4}
%\caption{Analyzing images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.97$. }
%\customlabel{fig:data2}{4}
%\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{A}}
%\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{B}}
%\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{C}}
%\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{D}}
%\end{figure*}

@@

>>>>

\begin{figure}[t]
\includegraphics{fig1}
\caption{Caricature illustrating the tasks of image registration and temporal ordering. (A) Images of ``samples'', each in a different orientation and a different stage of development. (B) Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
\end{figure}

@@

\begin{figure}[t]
\includegraphics{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. (A) Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in A: the differences between a pair of angles in B is the same as the pairwise angle in A. (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \SI), so that close data points are connected by darker (``stronger'') edges. (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
%\label{fig:schematics}
\customlabel{fig:schematics}{2}
\customlabel{subfig:synch1}{\ref{fig:schematics}A}
\customlabel{subfig:synch2}{\ref{fig:schematics}B}
\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
\end{figure}

@@

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.75cm}{
{ \textsf{ 
\begin{tabular}{ |p{1.25cm} | p{1.5cm}|  p{1.75cm}| }
\hline
Movie Index & Average Angle Error & Rank Correlation Coefficient \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\ 
2 & 4.98$^{\circ}$ & 0.9994 \\ 
3 & 6.47$^{\circ}$ & 0.9795 \\ 
4 & 7.03$^{\circ}$ & 0.9953 \\ 
5 & 3.04$^{\circ}$ & 0.9959 \\ 
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies. }
\end{figure*}

@@

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images. 
%% 
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
%
In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

\begin{figure*}[t]
\begin{minipage}[c]{13.75cm}
\vspace{0cm}
\raisebox{0.5cm}{\textsf{A}}
\includegraphics[width=13.5cm]{zebrafish_scrambled}

\raisebox{0.5cm}{\textsf{B}}
\includegraphics[width=13.5cm]{zebrafish_ordered}
\end{minipage}
%
\hfill
%
\begin{minipage}[c]{3.5cm}

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.15cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. Registered and ordered zebrafish data.}
\end{figure*}

To illustrate the broad applicability of our approach, we also applied our algorithm to a time-lapse movie of zebrafish embryogenesis. 
%
The original movie is available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \cite{karlstrom1996flipbook}. 
%
We analyzed 120 frames from this movie which span the epiboly stage.

As in the live imaging {\em Drosophila} example, the frames were randomly rotated and shuffled. 
%
We then used vector diffusion maps to register and order the frames. 
%
The results are shown in \fig~\ref{fig:zebrafish}. 
%
The recovered rotations and order are consistent with the expected developmental dynamics. 
%
Quantitatively, the rank correlation coefficient for this data set is $0.9935$, and the average error in the recovered angle is $4.23^\circ$.

@@

\raisebox{0.5cm}{\textsf{C}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
\end{minipage}
%
\hfill
%
\raisebox{3.25cm}{\textsf{D}}
\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
%
\caption{Analyzing images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.97$. }
\end{figure*}

We have analyzed how our algorithm performs on a model data set, where all images come from a single embryo. 
%
In practice, each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting an image from a random live imaging experiment at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.

@@

@@

\begin{figure*}[t]
\begin{minipage}[c]{9cm}
\vspace{0cm}

\raisebox{6.75cm}{\textsf{A}}
\includegraphics[width=8cm]{wing_disc_example}
\end{minipage}
%
\begin{minipage}[c]{8cm}
\caption{Registered and ordered three-dimensional wing-disc data}
\end{minipage}

\raisebox{0.7cm}{\textsf{B}}
\includegraphics[width=\textwidth]{wing_disc_ordered}
